---
title: "Replication, Estimator Performance: homogenous treatment effect model"
author: "[Christophe Bruneel-Zupanc](https://www.cbruneel.com/)"
date: "Last modified: 2024-07-22"
output:
  rmarkdown::html_document:
    toc: true
    toc_depth: 2
  rmarkdown::html_vignette:
    highlight: monochrome
    toc: true
  pdf_document:
    highlight: monochrome
bibliography: semiIVreg.bib
biblio-style: "apalike"
vignette: >
  %\VignetteKeywords{instrumental variables, semi-IVs, causal inference}
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{Replication, Estimator Performance: homogenous treatment effect model}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", echo = TRUE, fig.retina = 2#,#fig.height=4, fig.width=5, fig.align='center'
)
if (!requireNamespace("ivreg", quietly = TRUE)) {
  install.packages("ivreg")
}
library(ivreg); library(semiIVreg)
options(scipen=8)
```




The code below provides the code to run to replicate the Homogenous treatment effect simulations of @bruneel2024. 
For an exact replication, increase the number of Monte Carlo simulations to 500 (here, `N_simul=50` for pace). 


## Monte Carlo Simulation functions for model with Homogenous TE

```{r mc} 

# ----------------------------------
# 1. Monte-Carlo Simulation function
# ----------------------------------
MC_fun = function(N_simul, N, param) {

  param_error = param[[1]];
  param_Z = param[[2]];
  param_p = param[[3]]; param_y0 = param[[4]]; param_y1 = param[[5]]
  param_genX = param[[6]];
  model_type = param[[7]]

  # 1. Monte Carlo Simulation
  RES = list(); MTR = list(); MTRh = list()
  k=1

  while(k <= N_simul) {
    # Simulate:
    data = simul_data(N, model_type, param_y0, param_y1, param_p, param_Z, param_genX, param_error)

    # i) Naive OLS:
    naive_ols = lm(y ~ d + I((1-d)*w0) + I(d*w1), data); #summary(naive_ols)
    ols_est = naive_ols$coefficients

    # ii) Semi-IV 2SLS
    fstage = lm(d~w0+w1, data); summary(fstage)
    data$Phat = fitted(fstage)
    semiiv2sls = lm(y ~ Phat + I((1-Phat)*w0) + I(Phat*w1), data); #summary(semiiv2sls)
    semiiv2sls_est = semiiv2sls$coefficients

    # iii) Semi-IV 2SLS-bis - probit fstage
    fstage = glm(d~w0+w1, family=binomial(link="probit"), data=data)
    data$Phat = fitted(fstage)
    semiiv2sls_probit = lm(y ~ Phat + I((1-Phat)*w0) + I(Phat*w1), data); #summary(semiiv2sls)
    semiiv2sls_est_probit = semiiv2sls_probit$coefficients

    # iv) Semi-IV general
    semiiv = semiivreg(y ~ d | w0 | w1, ref_indiv = data.frame(w0=0, w1=0), data=data,
                       trimming_value = c(0.01, 0.99),
                       bw0 = 0.1, bw1=0.1, bw_y0=0.1, bw_y1=0.1,
                       fast_robinson2 = TRUE, plotting=FALSE)
    EY0 = mean(semiiv$data$RES$mtr0, na.rm=TRUE); EY1 = mean(semiiv$data$RES$mtr1, na.rm=TRUE); ATE = mean(semiiv$data$RES$mte, na.rm=TRUE)
    semiiv_est = c(EY0, ATE, semiiv$estimate$est0$coefficients, semiiv$estimate$est1$coefficients)
    MTR[[k]] = semiiv$data$RES;

    # v) Semi-IV homogenous:
    semiivh = semiivreg(y ~ d | w0 | w1, ref_indiv = data.frame(w0=0, w1=0), data=data, est_method="homogenous", plotting=FALSE)
    semiivh_est = semiivh$estimate$est[1:4, 2]
    MTRh[[k]] = semiivh$data$RES

    # vi) IVreg:
    iv = ivreg(y~d|w0+w1, data=data); #summary(iv)
    iv_est = c(iv$coefficients, NA, NA)

    # Save:
    save = rbind(ols_est, iv_est, semiiv2sls_est, semiiv2sls_est_probit, semiiv_est, semiivh_est)
    save = data.frame(save); names(save) = c("mu0", "mu1mu0", "delta0", "delta1")
    save$k = k
    save$method = c("ols", "iv", "semi2sls", "semiiv2sls_probit", "semiiv", "semiivh")

    RES[[k]] = save

    k = k+1;
    if(k%%10 == 0) { print(k)}
  }


  res_mc = do.call(rbind, RES);
  MTR = do.call(rbind, MTR)
  MTRh = do.call(rbind, MTRh)
  res = list(res_mc, MTR, MTRh)
  return(res)
}


```







## Baseline model with semi-IVs
### Specification 1

Adjust the Number of observations (`N`) and Number of Monte Carlo simulations (`N_simul`) accordingly. 

```{r model1} 

# ------------------------------------------------
# 2. Simulations with homogenous Treatment Effects - Baseline Model
# ------------------------------------------------
# Remark: param_genX are irrelevant in all simulations because I set the coefficients such that the covariates have no effect on anything;
N = 10000; N_simul=50;

# (i) Model
model_type = "homogenous"
param_error = c(1, 1, 0.6) # var_u, var_v, cov_uv # if homogenous
param_Z = c(0, 0, 0, 0, 1, 1, 0.5) # meanW0 state0, meanW1 state0, meanW0 state1, meanW1 state1, varW0, varW1, covW0W1
param_p = c(0, -0.7, 0.7, 0, 0, 0) # constant, alphaW0, alphaW1, alphaW0W1, effect of state, effect of parent educ
param_y0 = c(3.2, 0.8, 0, 0) # intercept, effect of Wd, effect of state, effect of parent educ;
param_y1 = c(3.2+0.4, 0.5, 0, 0) # the +0.2 = Average treatment effect; effect of W1, effect of state, effect of parent educ;
param_genX = c(0.4, 0, 2) # probability state=1 (instead of 0), mean_parenteduc, sd_parenteduc (parenteduc drawn as continuous)
# Remark: mean_V and the constant in the probit are playing the same role; normalize one to zero.
param = list(param_error, param_Z, param_p, param_y0, param_y1, param_genX, model_type)

# (ii) Monte Carlo:
res = MC_fun(N_simul, N, param)
res_mc = res[[1]]; MTR = res[[2]]; MTRh = res[[3]]

```




**Results of the estimation.** Specification 1 in the Table of the Online Appendix

```{r model1table} 
# (iii) Table of coefficients
# Table:
dmc = data.table(res_mc);
dmc[, list(mean_mu0=mean(mu0), sd_mu0=sd(mu0), mean_mu1mu0=mean(mu1mu0), sd_mu1mu0=sd(mu1mu0),
           mean_delta0=mean(delta0), sd_delta0=sd(delta0), mean_delta1=mean(delta1), sd_delta1=sd(delta1)), by=method]
```


**Plot.** Corresponding estimated MTRs:
```{r model1mtr, warning=FALSE} 

# (iv) Plot MTR
# a) Compute the true MTR:
seq_p = seq(0, 1, by=0.001);
w0 = 0; w1 = 0;
sigma_V2 = param_error[2]
covUV = param_error[3]
ku0 = covUV/sigma_V2*(qnorm(seq_p, mean=0, sd=sqrt(sigma_V2)) - 0) # 0 = mean(V)
ku1 = ku0 # 0 = mean(V)
true_mtr0 = param_y0[1] + param_y0[2]*w0 + ku0
true_mtr1 = param_y1[1] + param_y1[2]*w1 + ku1
true_mte = true_mtr1 - true_mtr0

newdata = data.frame(Ud=seq_p, w0=0, w1=0)
newdata$true_mte = true_mte; newdata$true_mtr1 = true_mtr1; newdata$true_mtr0 = true_mtr0


# b) MTR plot:
# Graphical options: 

colMTE = "#db93a4"; colD0 = "#f7ba4a"; colD1 = "#9993db"
colMTEtrue = "#DE40AA"; colD0true = "#EE7000"; colD1true = "#663A96"

dat = data.table(MTRh);

conf_level = 0.05

CI = dat[, .(
  count0 = sum(!is.na(mtr0)),
  count1 = sum(!is.na(mtr1)),
  count_mte = sum(!is.na(mte)),
  mte = mean(mte, na.rm=TRUE),
  mte_lwr = quantile(mte, conf_level/2, na.rm=TRUE),
  mte_upr = quantile(mte, 1-conf_level/2, na.rm=TRUE),
  mtr0 = mean(mtr0, na.rm=TRUE),
  mtr0_lwr = quantile(mtr0, conf_level/2, na.rm=TRUE),
  mtr0_upr = quantile(mtr0, 1-conf_level/2, na.rm=TRUE),
  mtr1 = mean(mtr1, na.rm=TRUE),
  mtr1_lwr = quantile(mtr1, conf_level/2, na.rm=TRUE),
  mtr1_upr = quantile(mtr1, 1-conf_level/2, na.rm=TRUE)
), by = .(Phat, id)]

RES_plot = CI;

common_supp_plot = c(0, 1)
mtr_plot = mtr_plot_fun(RES_plot, common_supp_plot, conf_band=TRUE)
mtr_p = mtr_plot +
  geom_line(data=newdata, aes(x=Ud, y=true_mtr0), color=colD0true, linetype="dashed") +
  geom_line(data=newdata, aes(x=Ud, y=true_mtr1), color=colD1true, linetype="dashed")
mtr_p = mtr_p + theme_classic() +
  theme(
    legend.text = element_text(size = 11), # if needs rescaling
    legend.position = c(0.97, 0),
    legend.justification = c("right", "bottom"), #, # Align the legend's bottom-left corner
    legend.background = element_blank()
  ) +
  scale_color_manual(values = c("0" = "#f7ba4a", "1" = "#9993db"), labels = c("0" = "d=0", "1" = "d=1")) +
  scale_fill_manual(values = c("0" = "#f7ba4a", "1" = "#9993db"), labels = c("0" = "d=0", "1" = "d=1")) +
  labs(color = NULL, fill = NULL) +
  ylab(bquote("Marginal Treatment Responses " ~ (MTR[d]))) +
  xlab("Unobserved resistance to treatment, V") +
  scale_x_continuous(limits = c(0, 1), expand = expand_scale(add = c(0, 0.02))) #+
#ylim(2.5, 4.6)
mtr_p

```










### Specification 2: What if the semi-IVs are valid IVs? 
Just change the effect of the semi-IVs on their respective potential outcomes to zero, and rerun the previous code. 


```{r model2} 
param_y0 = c(3.2, 0, 0, 0) # intercept, effect of Wd, effect of state, effect of parent educ;
param_y1 = c(3.2+0.4, 0, 0, 0) # the +0.2 = Average treatment effect; effect of W1, effect of state, effect of parent educ;
param = list(param_error, param_Z, param_p, param_y0, param_y1, param_genX, model_type)
```
